---
title: "Notes for Predictive Modelling"
subtitle: "MSc in Big Data Analytics at Carlos III University of Madrid"
author: "Eduardo García Portugués"
date: "`r Sys.Date()`, v0.1"
knit: "bookdown::render_book"
documentclass: book
bibliography: PM-UC3M.bib
biblio-style: apalike
link-citations: yes
site: bookdown::bookdown_site
---

# Introduction {#intro}

<!--
(Move up once decided what options)
cover-image: no
description: "A subject within the MSc in Big Data Analytics at Carlos III University of Madrid (Spain)."
github-repo: egarpor/PM-UC3M
apple-touch-icon: "touch-icon.png"
apple-touch-icon-size: 120
favicon: "favicon.ico"
-->

```{block2, type = 'rmdcaution'}
The **animations** of these notes will not be displayed the first time they are browsed^[The reason is because they are hosted at `https` websites with auto-signed SSL certificates.]. See for example Figure \@ref(fig:mse). **To see them**, click on the caption's link *"Application also available [here](https://ec2-35-177-34-200.eu-west-2.compute.amazonaws.com/dist-mse/)"*. You will get a warning from your browser saying that *"Your connection is not private"*. Click in *"Advanced"* and **allow an exception** in your browser (I guarantee you I will not do anything evil!). The next time the animation will show up correctly within the notes.
```

## Course objectives and logistics {#intro-course}

The **software** employed in the course is *the* statistical language [`R`](https://cran.r-project.org/) and its most common IDE (Integrated Development Environment) nowadays, [`RStudio`](https://www.rstudio.com/products/rstudio/download/). A basic prior knowledge of both is assumed^[Among others: basic programming in `R`, ability to work with objects and data structures, ability to produce graphics, knowledge of the main statistical functions, ability to run scripts in `RStudio`.]. The appendix presents basic introductions to `RStudio` and `R` for those students lacking basic expertise on them.

The notes contain a substantial amount of snippets of code that are fully self-contained. Students are encouraged to bring their own laptops at the lessons to practice with the code.

The required packages for the course are:
```{r, echo = TRUE, cache = TRUE, warning = FALSE, message = FALSE, eval = FALSE}
# Install packages
install.packages(c("ks", "nor1mix", "KernSmooth", "manipulate", "locfit"))
```

The codes in the notes may assume that the packages have been loaded, so it is better to do it now:
```{r, echo = TRUE, cache = TRUE, warning = FALSE, message = FALSE}
# Load packages
library(ks)
library(nor1mix)
library(KernSmooth)
library(manipulate)
library(locfit)
```
The **Shiny interactive apps** on the notes can be downloaded and run locally. This allows in particular to examine their codes. Check [this GitHub repository](https://github.com/egarpor/ShinyServer) for the sources.

Each topic of this contains a mix of theoretical and practical exercises for **grading**. Groups of *two students* must choose *three exercises* in total (at least one theoretical and other practical) from Sections \@ref(dens-exercises) and \@ref(reg-exercises) and turn them in order to be graded. The group grade is weighted according to the difficulty of the exercises, which is given by the number of stars: easy ($\star$), medium ($\star\star$), and hard ($\star\star\star$). The final grade ($0-5$) is
$$
\frac{1}{3}\sum_{i=1}^n\frac{\mathrm{Score}_i}{5}(2+\star_i),
$$
where $\mathrm{Score}_i$ is the score ($0-5$) for the $i$-th exercise and $\star_i$ represents its number of stars ($1-3$). Deadline for submission is 2017-10-13.

## Background and notation {#intro-background}

## Main references and credits {#intro-credits}

Several great reference books have been used for preparing these notes. The following list details the sections in which each of them has been consulted:

- @Wasserman2004

In addition, these notes are possible due to the existence of these incredible pieces of software: @R-bookdown, @R-knitr, @R-rmarkdown, and @R-base.

The icons used in the notes were designed by [madebyoliver](http://www.flaticon.com/authors/madebyoliver), [freepik](http://www.flaticon.com/authors/freepik), and [roundicons](http://www.flaticon.com/authors/roundicons) from [Flaticon](http://www.flaticon.com/).

All material in these notes is licensed under [CC BY-NC-SA 4.0](https://creativecommons.org/licenses/by-nc-sa/4.0/).
